---
title: "Lab 4: "
author: "Maarten van der Velde"
output: html_notebook
---

# 1. Preparation
```{r}
library(mgcv)
library(itsadug)
library(plyr)
library(dplyr)
library(colorspace)
```



Load data
```{r}
download.file("http://www.jacolienvanrij.com/Courses/LOT2018/data/dat_subset.rda", "dat_subset.rda")
load("dat_subset.rda")

head(dat)
str(dat)
summary(dat$Time)
```

Convert to factors
```{r}
dat <- dat %>%
  mutate(StimType = as.factor(StimType))
```



## Nonlinear interaction using `ti()`

```{r}
m2 <- bam(Pupil ~ StimType + s(Time, by=StimType)
          + s(Time, Subject, bs='fs', m=1)
          + s(Time, Item, bs='fs', m=1), data=dat, discrete = TRUE)
summary(m2)
```


Include the predictor `wordLength` and the interaction between `Time` and `wordLength`.
```{r}
m3 <- bam(Pupil ~ StimType +
            s(wordLength, by = StimType) + 
            s(Time, by = StimType) +
            ti(wordLength, Time, by = StimType) +
            s(Time, Subject, bs='fs', m=1) +
            s(Time, Item, bs='fs', m=1), 
          data=dat, discrete = TRUE)
summary(m3)
```

## Question 1: nonlinear interactions

**Does the nonlinear interaction contribute to the model?**

Yes. The interaction is a signifant effect and the model explains more of the variance in the data.

## Visualising nonlinear interactions


```{r}
pvisgam(m3, view=c("Time", "wordLength"), select=4, color=terrain_hcl(50))
abline(v=0, lty=3)
```

```{r}
fvisgam(m3, view=c("Time", "wordLength"), rm.ranef=TRUE, color=terrain_hcl(50))
abline(v=0, lty=3)
```

## Question 2: visualising nonlinear interactions

**What is the difference between the two functions?**

`pvisgam` plots the partial effects, in this case we only see the effect of the time smooth for 'word' stimuli. It shows that the pupil size increases over time.

`fvisgam` plots the summed fitted effects (i.e., the actual model prediction). If not specified, it just chooses a level for each factor (here `StimType` is set to pseudoword). Random effects can be included or not (here they are excluded).


## Nonlinear interactions using `te()`

We can replace the separate smooths and `ti()` term with a single `te()` term.
For simplicity, ignore the `StimType` predictor for now.

```{r}
m4a <- bam(Pupil ~ s(Time) + s(wordLength) + ti(Time, wordLength)
          + s(Time, Subject, bs='fs', m=1)
          + s(Time, Item, bs='fs', m=1), data=dat, discrete = TRUE, nthreads = 3)

m4b <- bam(Pupil ~ te(Time, wordLength)
          + s(Time, Subject, bs='fs', m=1)
          + s(Time, Item, bs='fs', m=1), data=dat, discrete = TRUE, nthreads = 3)

summary(m4a)
summary(m4b)
```

Although the models are comparable, they are not identical, as the plotted estimates show.
The `te()` version of the model is noticeably smoother, since it has fewer degrees of freedom to work with. In this case that might actually be a good thing, since the less constrained model predicts an effect of word length on pupil size before the word has even appeared, which should not be possible.

```{r}
par(mfrow=c(1,2))

fvisgam(m4a, view=c("Time", "wordLength"), rm.ranef=TRUE, color=terrain_hcl(50))
abline(v=0, lty=3)

fvisgam(m4b, view=c("Time", "wordLength"), rm.ranef=TRUE, color=terrain_hcl(50))
abline(v=0, lty=3)
```

## Question 3: `ti()` vs `te()`

**Translate `StimType + te(Time, wordLength, by=StimType)` to a comparable model formula with `ti` and `s`.**

`StimType + s(Time, by = StimType) + s(wordLength, by = StimType) + ti(Time, wordLength, by StimType)`

## Nonlinear interaction using `s()`

`s()` is used when the interacting predictors are isotropic (measured on the same scale and dimension).

Include a nonlinear interaction between gaze x and y coordinates:
```{r}
m5 <- bam(Pupil ~ StimType + s(Time, by=StimType) + s(wordLength)
          + ti(Time, wordLength)
          + s(median_GAZE_X, median_GAZE_Y)
          + s(Time, Subject, bs='fs', m=1)
          + s(Time, Item, bs='fs', m=1), data=dat, discrete = TRUE, nthreads = 4)

summary(m5)
```

Inspect the effect of gaze position on pupil size by visualising the partial effect:
```{r}
pvisgam(m5, view=c("median_GAZE_X", "median_GAZE_Y"), select=5, color=terrain_hcl(50))
points(dat$median_GAZE_X, dat$median_GAZE_Y)
```

## Question 4: Gaze position

**Describe the effect of gaze position on pupil size.**

It is not clear what the coordinate system is (is (0,0) the start of the word, the center of the screen, or something else?). Nevertheless we can see a fairly strong effect of the y-coordinate on pupil size: as participants look towards the top of the screen, their pupil becomes smaller. There is possibly also a weak interaction with the x-coordinate, which is especially visible when participants look towards the right of the screen.

**What is your conclusion? Do you see any problems?**

Aside from the issue that we don't know what the coordinates mean, the variation in pupil size is extremely big and occurs largely in one direction (vertically), which is suspicious.

