---
title: "Lab 5: Logistic Regression"
author: "Maarten van der Velde"
output: html_notebook
---

# 1. Preparation
```{r}
library(mgcv)
library(itsadug)
library(plyr)
library(dplyr)
library(colorspace)
```

Download data
```{r}
download.file("http://www.jacolienvanrij.com/Courses/LOT2018/data/dat_gaze.rda", "dat_gaze.rda")
load("dat_gaze.rda")

head(dat)
str(dat)
summary(dat$Time)
```

Select 2000 ms window:
```{r}
dat <- droplevels(dat[dat$Time > -500 & dat$Time < 2000,])
```


## Question 1: Relation proportion and logit

**How do positive logit values relate to proportions**

Proportions greater than 0.5 have a positive logit value.

**What value does logit value 0 translate to?**

0.5

**Advantage of logit scale over proportion scale?**

It is not truncated on either side, unlike a proportion scale, which is limited at 0 and 1.


# 2. Visualisation
```{r}
avg <- ddply(dat, c("Time", "verb_type"), summarise,
      IA_1 = sum(IA_1_C)/ sum(NSamples),
      IA_2 = sum(IA_2_C)/ sum(NSamples),
      IA_3 = sum(IA_3_C)/ sum(NSamples),
      IA_4 = sum(IA_4_C)/ sum(NSamples))

emptyPlot(c(0,2000), c(0,.5), v0=0,
          xlab="Time (ms)", ylab="Proportion",
          main="Gaze during sentence\nRestricting Verb")
with(avg[avg$verb_type=="Restricting",], lines(Time, IA_1, col=1, lwd=2) )
with(avg[avg$verb_type=="Restricting",], lines(Time, IA_2, col=2, lwd=2) )
with(avg[avg$verb_type=="Restricting",], lines(Time, IA_3, col=3, lwd=2) )
with(avg[avg$verb_type=="Restricting",], lines(Time, IA_4, col=4, lwd=2) )

legend('topleft', legend=c("Target"), col=1, lwd=2, bty='n')


emptyPlot(c(0,2000), c(0,.5), v0=0,
          xlab="Time (ms)", ylab="Proportion",
          main="Gaze during sentence\nNonRestricting Verb")
with(avg[avg$verb_type=="NonRestricting",], lines(Time, IA_1, col=1, lwd=2) )
with(avg[avg$verb_type=="NonRestricting",], lines(Time, IA_2, col=2, lwd=2) )
with(avg[avg$verb_type=="NonRestricting",], lines(Time, IA_3, col=3, lwd=2) )
with(avg[avg$verb_type=="NonRestricting",], lines(Time, IA_4, col=4, lwd=2) )


legend('topleft', legend=c("Target"), col=1, lwd=2, bty='n')
```

# 3. Start analysis

```{r}
dat$Target <- with(dat, cbind(IA_1_C, NSamples)) # make 2d column (why?!?) to store both area of interest 1 fixation count (IA_1_C) and the total number of samples in that time step.
dat$Distractor2 <- with(dat, cbind(IA_2_C, NSamples))

dat$verb_type <- as.factor(dat$verb_type) # gams expects factor
```

```{r}
m1 <- bam(Target ~ verb_type + s(Time, by=verb_type), data=dat, family='binomial')
```
Note that this gives a warning because the first part of the data is all zeroes (see the plots above). People cannot look anywhere in the first 200ms; they're still moving their eyes.

One solution: trim the first part off the timeseries:
```{r}
m1 <- bam(Target ~ verb_type + s(Time, by=verb_type), data=dat[dat$Time > 200,], family='binomial')
summary(m1)
```

Interpretation: the negative intercept (remember: it's on a logit scale) means that the average proportion of looks at the target is below 0.5. You can transform it back to a proportion for interpretation using `plogis`.
Similarly, when the verb is restricting, the proportion of looks to the target is even lower.

The edf values of the smooths are very close to their maximum, so it makes sense to allow more edf, by setting k = "some higher value".

```{r}
plot_smooth(m1, view = "Time", plot_all = "verb_type")
```

(Be aware that the confidence intervals are very small, indicating that the model is overconfident.)

This plot is on a logit scale, we can transform it back to a proportion scale using the `plogis` transformation in R.
```{r}
plot_smooth(m1, view = "Time", plot_all = "verb_type", transform = plogis)
```

Look at the residuals. Note that we would not expect normal residuals since the data are binomial.

```{r}
qqnorm(resid(m1))
qqline(resid(m1))
```

Are there autocorrelations in the residuals?

```{r}
acf(resid(m1))
```

Yes! Not surprising since we don't account for random effects in the model.




